\LARGE\singlespacing{\textbf{Supplementary Information} \\ \medskip Estimating \emph{Mycobacterium} tuberculosis transmission in a South African clinic: Spatiotemporal model based on person movements}

### 2024-11-13 doc/paper/supplements.tex

We build upon the Wells-Riley model\cite{Riley1961Book}, which is an established epidemiological model to estimate the risk of infection. The model estimates the risk (probability) of airborne indoor transmission $P$ using the following equation: 
\begin{align}
    P = \frac{C}{S} = 1 - \exp \left(-\frac{Ipqt}{Q}\right),
\end{align}
where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (s), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).

@@

The extension by Rudnick and Milton still assumes a well-mixed airspace. We relax this assumption by modelling spatial variation in the concentration of infectious quanta. We then compute the risk of infection for susceptible individuals using the Wells-Riley equation considering their cumulative exposure to infectious quanta over space and time as 
\begin{equation}\label{eq:spattemp-P}
    P = \sum_s \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p,
\end{equation}
where $N$ is the quanta concentration in the indoor space area $c$ at time $t$ (in quanta/$m^3$), $\mathbb{I}$ is a binary variable indicating whether the susceptible individual was in $c$ at $t$, and $p$ is the breathing rate.

@@

We divide the airspace into a 2-dimensional grid with cells $c = 1, \dots, C$, where $(x_c, y_c)$ refer to the centre of cell $c$. The grid provides a discrete approximation to the continuously varying quanta concentration in the indoor space, with the number of grid cells $C = n_x \cdot n_y$ determining the spatial resolution. We denote the following variables and modelling parameters:
\begin{itemize}
    \item $N_{c,t}$: quanta concentration in airspace unit $c = 1, \dots, C$ at time $t = 1, \dots T$ (in seconds)
    \item $I_{c,t}$: number of infectious individuals in airspace unit $c$ at time $t$
    \item $q$: quanta generation rate (quanta s$^{-1}$)
    \item $D$: diffusion constant (m$^2$ s$^{-1}$)
    \item $AER$: air change rate (air changes s$^{-1}$)
    \item $V$: volume of the airspace (m$^3$)
    \item $CO^{\text{I}}$: CO$_2$ mixing ratio in indoor air (parts per million, ppm)
    \item $CO^{\text{O}}$: CO$_2$ mixing ratio in outdoor air (ppm)
    \item $G$: average CO$_2$ generation rate per person (L$\cdot$min$^{-1}\cdot$person$^{-1}$)
    \item $n_t$: number of individuals in the airspace at time $t$
    \item $\lambda$: bacterial inactivation rate (quanta s$^{-1}$)
    \item $k$: gravitational settling rate (quanta s$^{-1}$)
\end{itemize}

@@

\Cref{fig:toy-example} shows the spatiotemporal quanta concentration. The quanta concentration increases over time and is higher near the infectious individual in the room's centre (\Cref{fig:toy-example}a, top panel). After the individual has left the room, the quanta diffuses within minutes and most quanta is removed by 10\,am (\Cref{fig:toy-example}a, bottom panel) through outdoor air exchange. Spatial variation in the quanta concentration peaks at 9\,am, right before the infectious individual leaves the space, and then it takes about 15\,min until the a well-mixed airspace has been reached through quanta diffusion (\Cref{fig:toy-example}b).

\begin{figure}[!htpb]
    \centering
    \includegraphics{tests/stm_v2-toy_example.png}
    \includegraphics{tests/stm_v2-toy_example-B.png}
    \caption[Example: Spatiotemporal quanta concentration in an indoor space]{\textbf{Example: Spatiotemporal quanta concentration in an indoor space between 8\,am and 10\,am following the generation of quanta by an infectious individual between 8\,am and 9\,am in the room's centre}. \textbf{(a)}~Spatial quanta concentration at selected time points and \textbf{(b)}~quanta concentration over time per cell of the indoor space.}
    \label{fig:toy-example}
\end{figure}

@@

Previous studies estimated the quanta generation rate of \emph{Mtb} using the Wells-Riley model\cite{Andrews2014JID,Riley1962ARRD,Escombe2008PLoSMed,Nardell1991ARRD,Dharmadhikari2012AJRCCM}. Andrews et al.\cite{Andrews2014JID} reported an estimate of 0.89\,(quanta h$^{-1}$) (range from 0.44 to 5.69), close to the early estimate provided by Riley et al.\cite{Riley1962ARRD} with 1.25\,quanta h$^{-1}$. On the other end, Escombe et al.\cite{Escombe2008PLoSMed} arrived at an average estimate of 8.2\,quanta h$^{-1}$, which, however, could be as high as 226\,quanta h$^{-1}$. Lower rates were observed in non-multidrug-resistant (MDR), smear-negative patients and considerably higher ones in MDR, smear-positive TB patients\cite{Escombe2008PLoSMed}. Dharmadhikari et al.\cite{Dharmadhikari2012AJRCCM} also estimated a high average rate of 138\,quanta h$^{-1}$ in MDR TB patients, and Nardell et al.\cite{Nardell1991ARRD} reported an estimate of 13\,quanta h$^{-1}$ for a single index case. The wide range of estimates within and between studies suggest that the emitted quanta varies considerably with the infectiousness of \emph{Mtb} patients\cite{Wurie2016BMJ}. To consider such variation, Mikszewski et al.\cite{Mikszewski2021GF} used the bacterial load distribution in sputum to estimate the quanta generation rate with the approach developed by Buonanno et al.\cite{Buonanno2020EI}. We have recently adopted this approach to derive quanta generation rate distributions for different activity levels in a school setting\cite{Banholzer2024PGPH}. Here, we follow the same approach to derive a quanta generation rate distribution for sitting and walking activities, assuming 80\% breathing and 20\% speaking during both activities. The resulting prior distributions are shown in \Cref{fig:quanta-distribution} along with the estimates from the literature. The quanta generation rate follows a lognormal prior distribution with median 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking. The activity is determined based on whether the the person had moved by more or less than 0.25\,m (spatial distance between two recorded tracks of the same person).

@@

Adams (1993)\cite{Adams1993} reports breathing rates for different activity levels. We use the average breathing rate for the following activity levels: sitting with $p = 0.51$\,m$^3$ h$^{-1}$ and walking (slowly at 2.5 mph) with $p = 1.33$\,m$^3$ h$^{-1}$.

@@

% \begin{figure}[!htpb]
%     \centering
%     \includegraphics{results/data/diagnosed-tb-patients.png}
%     \caption{Number of diagnosed TB patients (dark bars) and those suspected of TB (light bars) following symptom screening. Modeled study days in color, other clinic days in October and November in grey.}
%     \label{fig:diagnosed-tb-patients}
% \end{figure}

@@

% \begin{figure}
%     \centering
%     \includegraphics{results/modeling/patient-risk-focus-plot.png}
%     \caption{Mean and 80\%-CrI (uncertainty) of the risk of infection per clinic visit. Highlighted in red is a visit from a clinic attendee with moderate risk of infection and low estimated uncertainty.}
%     \label{fig:patient-risk-focus}
% \end{figure}

>>>>

We build upon the Wells-Riley model\cite{Riley1961Book}, which is an established epidemiological model to estimate the risk of infection. The model estimates the risk (probability) of airborne indoor transmission $P$ using the following equation: 
\begin{align}
    P = \frac{C}{S} = 1 - \exp \left(-\frac{Ipqt}{Q}\right),
\end{align}
where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$\,h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta\,h$^{-1}$), $t$ is the exposure time (s), and $Q$ is the ventilation rate (m$^3$\,h$^{-1}$).

@@

The extension by Rudnick and Milton still assumes a well-mixed airspace. We relax this assumption by modelling spatial variation in the concentration of infectious quanta. We then compute the risk of infection for susceptible individuals using the Wells-Riley equation considering their cumulative exposure to infectious quanta over space and time as 
\begin{equation}\label{eq:spattemp-P}
    P = \sum_s \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p,
\end{equation}
where $N$ is the quanta concentration in the indoor space area $c$ at time $t$ (in quanta\,$m^3$), $\mathbb{I}$ is a binary variable indicating whether the susceptible individual was in $c$ at $t$, and $p$ is the breathing rate (in m$^3$\,h$^{-1}$).

@@

We divide the airspace into a 2-dimensional grid with cells $c = 1, \dots, C$, where $(x_c, y_c)$ refer to the centre of cell $c$. The grid provides a discrete approximation to the continuously varying quanta concentration in the indoor space, with the number of grid cells $C = n_x \cdot n_y$ determining the spatial resolution. We denote the following variables and modelling parameters:
\begin{itemize}
    \item $N_{c,t}$: quanta concentration in airspace unit $c = 1, \dots, C$ at time $t = 1, \dots T$ (in seconds)
    \item $I_{c,t}$: number of infectious individuals in airspace unit $c$ at time $t$
    \item $q$: quanta generation rate (quanta s$^{-1}$)
    \item $D$: diffusion constant (m$^2$ s$^{-1}$)
    \item $AER$: air change rate (air changes s$^{-1}$)
    \item $V$: volume of the airspace (m$^3$)
    \item $CO^{\text{I}}$: CO$_2$ mixing ratio in indoor air (parts per million, ppm)
    \item $CO^{\text{O}}$: CO$_2$ mixing ratio in outdoor air (ppm)
    \item $G$: average CO$_2$ generation rate per person (L$\cdot$min$^{-1}\cdot$person$^{-1}$)
    \item $n_t$: number of individuals in the airspace at time $t$
    \item $\lambda$: bacterial inactivation rate (quanta s$^{-1}$)
    \item $k$: gravitational settling rate (quanta s$^{-1}$)
    \item $p$: breathing rate (m$^3$\,h$^{-1}$)
\end{itemize}

@@

\Cref{fig:toy-example} shows the spatiotemporal quanta concentration. The quanta concentration increases over time and is higher near the infectious individual in the room's centre (\Cref{fig:toy-example}a, top panel). After the individual has left the room, the quanta diffuses within minutes and most quanta is removed by 10\,am (\Cref{fig:toy-example}a, bottom panel) through outdoor air exchange. Spatial variation in the quanta concentration peaks at 9\,am, right before the infectious individual leaves the space, and then it takes about 15\,min until a well-mixed airspace has been reached through quanta diffusion (\Cref{fig:toy-example}b).

\begin{figure}[!htpb]
    \centering
    \includegraphics{tests/stm_v2-toy_example.png}
    \includegraphics{tests/stm_v2-toy_example-B.png}
    \caption[Example: Spatiotemporal quanta concentration in an indoor space]{\textbf{Example: Spatiotemporal quanta concentration in an indoor space between 8\,am and 10\,am following the generation of quanta by an infectious individual between 8\,am and 9\,am in the room's centre}. \textbf{(a)}~Spatial quanta concentration at selected time points and \textbf{(b)}~quanta concentration over time for the 800~cells into which the indoor space was divided.}
    \label{fig:toy-example}
\end{figure}

@@

Previous studies estimated the quanta generation rate of \emph{Mtb} using the Wells-Riley model\cite{Andrews2014JID,Riley1962ARRD,Escombe2008PLoSMed,Nardell1991ARRD,Dharmadhikari2012AJRCCM}. Andrews et al.\cite{Andrews2014JID} reported an estimate of 0.89\,(quanta h$^{-1}$) (range from 0.44 to 5.69), close to the early estimate provided by Riley et al.\cite{Riley1962ARRD} with 1.25\,quanta h$^{-1}$. On the other end, Escombe et al.\cite{Escombe2008PLoSMed} arrived at an average estimate of 8.2\,quanta h$^{-1}$, which, however, could be as high as 226\,quanta h$^{-1}$. Lower rates were observed in non-multidrug-resistant (MDR), smear-negative patients and considerably higher ones in MDR, smear-positive TB patients\cite{Escombe2008PLoSMed}. Dharmadhikari et al.\cite{Dharmadhikari2012AJRCCM} also estimated a high average rate of 138\,quanta h$^{-1}$ in MDR TB patients, and Nardell et al.\cite{Nardell1991ARRD} reported an estimate of 13\,quanta h$^{-1}$ for a single index case. The wide range of estimates within and between studies suggest that the emitted quanta varies considerably with the infectiousness of \emph{Mtb} patients\cite{Wurie2016BMJ}. To consider such variation, Mikszewski et al.\cite{Mikszewski2021GF} used the bacterial load distribution in sputum to estimate the quanta generation rate with the approach developed by Buonanno et al.\cite{Buonanno2020EI}. We have recently adopted this approach to derive quanta generation rate distributions for different activity levels in a school setting\cite{Banholzer2024PGPH}. Here, we follow the same approach to derive a quanta generation rate distribution for sitting and walking activities, assuming 80\% breathing and 20\% speaking during both activities. The resulting prior distributions are shown in \Cref{fig:quanta-distribution} along with the estimates from the literature. The quanta generation rate follows a lognormal prior distribution with median 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking. The activity is determined based on whether the person had moved by more or less than 0.25\,m (spatial distance between two recorded tracks of the same person).

@@

Adams (1993)\cite{Adams1993} reports breathing rates for different activity levels. We use the average breathing rate for the following activity levels: sitting with $p = 0.51$\,m$^3$\,h$^{-1}$ and walking (slowly at 2.5 mph) with $p = 1.33$\,m$^3$\,h$^{-1}$.

@@

@@
